Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-11 TOMS_Paper/assembly.tex
Rewrite some introduction

Achieving high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Also, traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. This has led to development of a number of higher-level approaches to optimize local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of the problem. In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.

Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework. While clearly separating the mathematical domain, which remains captured in the higher-level form compiler, from the optimization process, COFFEE also aims to be platform-agnostic. The code transformations occur on an intermediate representation of the assembly code, which is ultimately translated into platform-specific code. Domain knowledge is exploited in two ways: for simplifying the implementation of a broad range of code transformations, and, obviously, to make them extremely effective. Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input. For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process. Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only. We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.

In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation. We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization. The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become sub-steps of code specialization; on the other hand, generalized loop-invariant code motion is a step of the expression rewriting process. More importantly, we complement and generalize our previous work with the following contributions.

Expression rewriting is based on a formal set of rewrite rules. Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of arithmetic operators to expose ``hidden'' loop-invariant sub-expressions. Secondly, we show how to make use of domain knowledge to avoid computation over zero-valued regions in vector-valued basis functions arrays, while preserving code vectorizability. These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable by using FEniCS' built-in optimizations, presented in~\cite{quadrature1}.

At code specialization time, transformations are applied to maximize the exploitation of the underlying platform's resources, e.g. SIMD lanes. On top of the work in~\cite{Luporini}, we provide a number of contributions. Firstly, we show the benefit of vector-expansion to achieve SIMD vectorization of otherwise scalar code. This is particularly useful in complex forms, like those based on hyperelasticity. Secondly, we answer an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix multiplies. BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods. However, we will show that in corner, yet important cases, especially in forms characterized by pre-multiplying functions and relatively high-order function spaces, a BLAS-based execution strategy can be a successful. Finally, we introduce a model-driven, dynamic autotuner that automatically and transparently compose the set of code transformations that are likely to maximize the performance of a given problem. The main challenge with the autotuner is to maintain, for any possible problem, the search space reasonably small, although comprising the most effective code variants, so that the overhead, which impacts the run-time, is negligible.

Expression rewriting and code specialization have been implemented in COFFEE and are fully operating. Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation in a number of forms of increasing complexity, including problems based on hyperelasticity operators. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare four sets of code variants, for each problem instance: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper. Notable performance improvements of 4) over 1), 2) and 3) are reported and detailed.

>>>>

Producing high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. Polyhedral-model-based source-to-source compilers, for instance~\cite{PLUTO}, mainly apply aggressive loop optimizations, such as tiling, but these are not particularly helpful in our context. This lack of suitable optimizing tools has led to the development of a number of higher-level approaches to maximize the performance of local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration. In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct set of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.

Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework~\cite{Luporini}. Besides separating the mathematical domain, captured by the form compiler at an higher level of abstraction, from the optimization process, COFFEE also aims to be platform-agnostic. The code transformations occur on an intermediate representation of the assembly operation, which is ultimately translated into platform-specific code. Domain knowledge is exploited in two ways: for simplifying the implementation of code transformations and to make them extremely effective. Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input. For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process. Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only. We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.

In this paper, we build on our previous work~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines based on quadrature representation. We argue that for complex, realistic forms, peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to ensure effective register utilization and SIMD vectorization. The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become steps of code specialization, whereas generalized loop-invariant code motion is performed during the expression rewriting stage. We complement and generalize our previous work with the following contributions.

Expression rewriting aims at reducing the computational intensity of local assembly kernels by rescheduling arithmetic operations based on a set of rewrite rules. Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage. While rewriting an assembly expression, domain knowledge is used in several flavours, for example to avoid redundant computation. Secondly, we propose an algorithm that, relying on symbolic execution, allows to restructure the assembly expression so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability. These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.

At code specialization time, transformations are applied to maximize the exploitation of the underlying platform's resources, e.g. SIMD lanes. On top of the work in~\cite{Luporini}, we provide a number of contributions. Firstly, we revisit the padding and data alignment transformation to account for the new code structures produced by expression rewriting. This is of fundamental importance to achieve effective SIMD vectorization. Secondly, we address an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix-matrix multiply routines. BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods. However, we will show that in corner, yet meaningful cases, especially in forms characterized by the presence of pre-multiplying functions and relatively high-order function spaces, a BLAS-based implementation can be successful. Finally, we introduce a model-driven, dynamic autotuner that transparently evaluates multiple sets of code transformations to determine the best optimization strategy for a given problem. The main challenge here is to build, for a generic form, a reasonably small search space that comprises most of the effective code variants.

Expression rewriting and code specialization have been implemented in COFFEE, which in turn is integrated with the Firedrake framework. Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on hyperelasticity. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare, for each test case, four sets of code variants: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper. Notable performance improvements of 4) over 1), 2) and 3) are reported and discussed.
